Ultra-compact, widely bandwidth-tunable filter using cascaded silicon photonic contra-directional couplers

### 2015-06-11 CacadedTunableFilter.tex
Abstract

\begin{abstract}
We demonstrate an integrated silicon tunable band-pass filter with low in-band ripples of 0.3 dB, a high contrast of 55 dB, and no free-spectral range. 
A record bandwidth tuning from 788 to 117 GHz was achieved.
\end{abstract}

@@

%Next-generation optical transmission systems require dynamic, efficient bandwidth allocation  
%Bandwidth tunable optical filters are essential components for optical communication applications such as reconfigurable optical wavelength-division multiplexing (WDM), signal processing, and dynamic bandwidth allocation \cite{DynamicBW}.
%These key devices are now needed in integrated photonic systems, in this case, on the silicon-on-insulator platform (SOI), to achieve the same functions as traditional discrete optical components such as diffractive grating spectrometers. 
Next-generation optical transmission systems applying flexible networking and the super-channel technique will require highly dynamic channel allocation to drastically increase the spectral efficiency and transmission capacity \cite{jinno2009spectrum, geisler2011demonstration}.
%over a few hundred gigahertz \cite{jinno2009spectrum}. 
%bandwidth
Tunable optical filters, reconfigurable in both center wavelength and bandwidth with scalability towards terahertz \cite{geisler2011demonstration}, are essential for these applications.
Such large bandwidth tunability is currently only available in bulky bench-top systems using diffractive grating spectrometers or liquid crystals.
Integrated solutions are desired for lower cost and power consumption.
In particular, silicon photonics  based on the submicron silicon-on-insulator platform allows for CMOS compatible mass fabrication, enabling low cost, high yield, and high-density chip-scale integration.
Existing solutions for tunable  filters on silicon include devices based on microring resonators \cite{DynamicBW, ong2013ultra} and Mach-Zehnder interferometers (MZIs). These devices have relatively small tunable bandwidth (less than 200 GHz) and small free spectral ranges (FSRs) (typically less than 10 nm), not suitable for high-capacity transmission applications.

Contra-directional couplers(contra-DCs) are grating-assisted add-drop filters \cite{shi2013siliconContraDC}. 
Analogous to waveguide Bragg gratings, the wavelength selectivity in contra-DCs are based on periodic dielectric perturbations. But instead of backreflections in the same waveguide, the selected wavelength in a contra-DC is dropped to another waveguide through contra-directional coupling.
This allows add-drop operation without the need of a circulator. 
Contra-DCs have merits of compactness, flat-top response, flexible filter design (e.g, through apodization), and near-infinite FSR (in the case of first-order gratings).
In particular, they allow for very high bandwidths (greater than 10 nm) and thus can support very-high-baud-rate super-channel signals \cite{jinno2009spectrum}.
In this paper, we demonstrate a broadband filter with a large tunability in both wavelength and channel bandwidth, using thermally controlled cascaded contra-DCs on a silicon chip. 
The filter has flat-top responses, low insertion loss, low in-band ripples, and high contrast between the pass-band and the stop-band.

@@

On top of each contra-DC sits a metal strip acting as a microheater for thermal tuning.  
The refractive index of silicon has a temperature dependence of $\frac{\diff{n_\text{Si}}}{\diff{T}}=$ 1.87$\times10^{-4}$K$^{-1}$ at room temperature for wavelengths around 1550 nm~\cite{frey2006temperature}.
Without electrical input, the optical responses of the two filters are identical (neglecting fabrication uniformities) resulting in a sharp transition from the pass-band to the stop-band. 
Applying independent electrical currents on the heaters, we can shift the spectra of the contra-DCs simultaneously or differentially for wavelength or bandwidth tuning.
In the ideal case, each contra-DC should be uniformly heated so that the apodization profile isn't disturbed as the temperature varies.

@@

@@

@@

%\section{Conclusion}
In summary, we have demonstrated a bandwidth tunable filter with a low insertion loss of less then 0.5 dB, low ripples of less than 0.3 dB, a large maximal bandwidth of greater than 750 GHz, and high contrast of 55 dB. 
Compared to other devices based on microring resonators and MZIs (see Table \ref{table:comparison}), it enables a much larger maximal passband and does not suffer from small FSRs. 
A large bandwidth tuning range over 670 GHz has been achieved, which, to the best of our knowledge, is the widest ever demonstrated on a silicon chip. 
This ultra-wide bandwidth tunability makes the device very attractive for next-generation ultrahigh baud-rate applications (e.g., high-capacity super-channel transmissions) and flexible optical networking.

\begin{table*}[t]
\caption{Recent results with multi-element on-chip silicon filters}
\begin{tabular}{cccccc}
    \hline
	Publication & Filter type & BWmax / FSR & Tunable BW & On-chip loss & Contrast \\
    \hline
    %Ibrahim (2011) &	Cascaded MRs &	0.4 GHz / 10 GHz &	0.6-2 GHz &	0.6 dB &	30 dB
    % \\
    %Alipour (2011) &	Cascaded MRs &	5 GHz / 650 Ghz &	0.8-5 GHz &	1.25-3.75 dB &	38 dB
    % \\
    Ding (2011) &	MRs + MZI &	55 GHz / 1 THz &	28-55 GHz &	3.6 dB &	30 dB
     \\
 	Orlandi (2012) &	MRs + MZI &	173 GHz / 200 GHz &	23-173 GHz &	0.46-1.06 dB &	15-34 dB
      \\
    %Luo (2012) &	Cascaded MRs &	100 GHz / 750 Ghz &	No &	0.3-0.6 dB &	50 dB  \\
    Ong (2013)\cite{ong2013ultra} &	Cascaded MRs &	125 GHz/ 0.9 THz &	11.6-125 GHz &	0.25 dB &	50-100 dB
      \\
    This work & Cascaded contra-DC &	778 GHz / Unlimited  &	117-778 GHz &	0.5 dB &	55 dB \\

>>>>

\begin{abstract}
%We demonstrate an integrated silicon tunable band-pass filter with low in-band ripples of 0.3 dB, a high contrast of 55 dB, and no free-spectral range. 
%A record bandwidth tuning from 788 to 117 GHz was achieved.

We demonstrate an integrated tunable band-pass filter continuously tuned from 788 GHz down to 117 GHz of 3 dB bandwidth with up to 3 nm of wavelength tunability at 1550 nm. The out of band contrast is up to 55 dB, the in band ripples are under 0.3 dB and the free spectral range is unlimited. This result was achieved using contra-directional couplers and micro-heaters on the silicon-on-insulator platform with a footprint of less than 0.007 mm$^2$.
\end{abstract}

@@

%Next-generation optical transmission systems require dynamic, efficient bandwidth allocation  
%Bandwidth tunable optical filters are essential components for optical communication applications such as reconfigurable optical wavelength-division multiplexing (WDM), signal processing, and dynamic bandwidth allocation \cite{DynamicBW}.
%These key devices are now needed in integrated photonic systems, in this case, on the silicon-on-insulator platform (SOI), to achieve the same functions as traditional discrete optical components such as diffractive grating spectrometers. 
Next-generation optical transmission systems applying flexible networking and the super-channel technique will require highly dynamic channel allocation to drastically increase the spectral efficiency and transmission capacity \cite{jinno2009spectrum, geisler2011demonstration}.
%over a few hundred gigahertz \cite{jinno2009spectrum}. 
%bandwidth
Tunable optical filters, reconfigurable in both center wavelength and bandwidth with scalability towards terahertz \cite{geisler2011demonstration}, are essential for these applications.
Such large bandwidth tunability is currently only available in bulky bench-top systems using diffractive grating spectrometers or liquid crystals.
Integrated solutions are desired for lower cost and power consumption.
In particular, silicon photonics  based on the sub-micron silicon-on-insulator platform allows for CMOS compatible mass fabrication, enabling low cost, high yield, and high-density chip-scale integration.
Existing solutions for tunable  filters on silicon include devices based on microring resonators \cite{DynamicBW, ong2013ultra} and Mach-Zehnder interferometers (MZIs). These devices have relatively small tunable bandwidth (less than 200 GHz) and small free spectral ranges (FSRs) (typically less than 10 nm), not suitable for high-capacity transmission applications.

Contra-directional couplers(contra-DCs) are grating-assisted add-drop filters \cite{shi2013siliconContraDC}. 
Analogous to waveguide Bragg gratings, the wavelength selectivity in contra-DCs are based on periodic dielectric perturbations. But instead of back-reflections in the same waveguide, the selected wavelength in a contra-DC is dropped to another waveguide through contra-directional coupling.
This allows add-drop operation without the need of a circulator. 
Contra-DCs have merits of compactness, flat-top response, flexible filter design (e.g, through apodization), and near-infinite FSR (in the case of first-order gratings).
In particular, they allow for very high bandwidths (greater than 10 nm) and thus can support very-high-baud-rate super-channel signals \cite{jinno2009spectrum}.
In this paper, we demonstrate a broadband filter with a large tunability in both wavelength and channel bandwidth, using thermally controlled cascaded contra-DCs on a silicon chip. 
The filter has flat-top responses, low insertion loss, low in-band ripples, and high contrast between the pass-band and the stop-band.

@@

On top of each contra-DC sits a metal strip acting as a micro-heater for thermal tuning.  
The refractive index of silicon has a temperature dependence of $\frac{\diff{n_\text{Si}}}{\diff{T}}=$ 1.87$\times10^{-4}$K$^{-1}$ at room temperature for wavelengths around 1550 nm~\cite{frey2006temperature}.
Without electrical input, the optical responses of the two filters are identical (neglecting fabrication uniformities) resulting in a sharp transition from the pass-band to the stop-band. 
Applying independent electrical currents on the heaters, we can shift the spectra of the contra-DCs simultaneously or differentially for wavelength or bandwidth tuning.
In the ideal case, each contra-DC should be uniformly heated so that the apodization profile isn't disturbed as the temperature varies.

@@

While only coupling apodization has been discussed so far, it is common in Bragg grating profiles to also adjust the local Bragg wavelength. This is done with a phase shift changing the local effective index or period. This design only uses coupling apodization in principle, but phase noise also appears during fabrication due to non-uniformity and sidewall roughness.
There are 3 main types of noise: high frequency (typically around 10 $\mu m$m\cite{simard2013characterization}), which is caused by sidewall roughness, decreases the sidelobe suppression; low frequency (larger than the device or 300 $\mu$m), which is caused by wafer thickness and etching variance, causes a linear chirp that increases the bandwidth; and coupling dependent, which is due to an index difference between low coupling and high coupling regions.

It is possible to easily add high frequency phase noise to the transfer matrix solution by adding random noise to the $\Delta\hat{\beta}$ propagation constants. The frequency distribution of the noise will thus match the quantization steps of the matrices, but since the response noise is dependent of both the amplitude and frequency of the sidewall roughness \cite{simard2011impact},we can match the simulation to the experiment by choosing the right noise amplitude.
Similarly, we can also add a variation to $\Delta\hat{\beta}$ which is dependent on the longitudinal position and the local coupling does not change as the temperature varies.

@@

We observe that the central wavelength at higher temperature is out of the original band, indicating thermal coupling between the two filters. Also, the side-lobes suppression ratio is lower for smaller bandwidth due to the signal only being affected by one filter. To solve both of these problems, next generation devices should use a isolation or larger spacing between filter elements (we used 12 \text{$\mu$}m center to center) and additional filtering stages.

@@

Table \ref{table:comparison} shows a comparison of this filter with other recent publications of tunable filters. No other tunable filter has such a large available bandwidth and FSR, making contra-DC unique in their application potential.

%\section{Conclusion}
In summary, we have demonstrated a bandwidth tunable filter with a low insertion loss of less then 0.5 dB, low ripples of less than 0.3 dB, a large maximal bandwidth of greater than 750 GHz, and high contrast of 55 dB. 
A large bandwidth tuning range over 670 GHz has been achieved, which, to the best of our knowledge, is the widest ever demonstrated on a silicon chip. 
This ultra-wide bandwidth tunability makes the device very attractive for next-generation ultrahigh baud-rate applications (e.g., high-capacity super-channel transmissions) and flexible optical networking.

\begin{table*}[t]
\caption{Recent results with multi-element on-chip silicon filters}
\begin{tabular}{cccccc}
    \hline
	Publication & Filter type & BWmax / FSR & Tunable BW & On-chip loss & Contrast \\
    \hline
    %Ibrahim (2011) &	Cascaded MRs &	0.4 GHz / 10 GHz &	0.6-2 GHz &	0.6 dB &	30 dB
    % \\
    %Alipour (2011) &	Cascaded MRs &	5 GHz / 650 Ghz &	0.8-5 GHz &	1.25-3.75 dB &	38 dB
    % \\
    Ding (2011)\cite{ding2011bandwidth} &	MRs + MZI &	55 GHz / 1 THz &	28-55 GHz &	3.6 dB &	30 dB
     \\
 	Orlandi (2012)\cite{orlandi2012reconfigurable} &	MRs + MZI &	173 GHz / 200 GHz &	23-173 GHz &	0.46-1.06 dB &	15-34 dB
      \\
    %Luo (2012) &	Cascaded MRs &	100 GHz / 750 Ghz &	No &	0.3-0.6 dB &	50 dB  \\
    Ong (2013)\cite{ong2013ultra} &	Cascaded MRs &	125 GHz/ 0.9 THz &	11.6-125 GHz &	0.25 dB &	50-100 dB
      \\
    This work & Cascaded contra-DC &	778 GHz / Unlimited  &	117-778 GHz &	0.5 dB &	55 dB \\
